---
title: "Time Series Project"
author: "Maya Elshweikhy - 900204233, Alaa Aoun - 900204512,and Karim AbouDaoud - 900212779"
date: "5/22/2024"
output: word_document
---
## Daily Climate Time Series Data Analysis 
- The data was obtained from kaggle and we used the mean temprature column for our analysis. 
- Data Source: https://www.kaggle.com/datasets/sumanthvrao/daily-climate-time-series-data/data

1. Reading the dataset
```{r}
setwd("~/Desktop/Spring 2024/Time Series/Project")
data<-read.csv("Final Data.csv")
my_data<-data[,2]
head(my_data)
```

```{r}
plot.ts(my_data)

```
This is the graph of our data. There doesn't seem a trend. We will now check for stationarity.

2. Testing for stationarity
```{r}
library(urca)
df1<-ur.df(my_data,type="none",lags=1)
summary(df1)
```
By looking at z.lag.1 we notice that the pr is 0.316 > 0.05, we fail to reject H_0. This means we will try to take difference and then check for stationarity again. 

3. Taking first difference

```{r}
dif_data1<-diff(my_data, diff=1)
```
After taking the first difference, we will test for stationarity again.
4. Checking for stationarity again
```{r}
df1new<-ur.df(dif_data1,type="none",lags=1)
summary(df1new)
```
When we did the test for stationarity again, the z.lag.1 gave pr of 2e-16 which is much less than 0.05. This means that we reject H_0. This implies that now our series is stationary. 

5. Checking for stationarity using a different method
```{r}
library(tseries)
df2<-adf.test(dif_data1,k=1)
df2
```
This output shows that p-value is 0.01<0.05. This implies that we reject H_0 and thus we can confirm that our series is stationary and we can proceed with our analysis.

6. Plotting ACF and PACF
```{r}
par(mfrow=c(1,2))
acf(dif_data1)
pacf(dif_data1)
```
It is noticeable that our ACF cuts off at lag 3 and PACF decays. We can say that it is MA(3). 

7. Modeling the data (MA(3))

```{r}
ma3<-arima(my_data, order=c(0,1,3))
ma3
```

**Model Equation:**
X_t= -0.2261 e_t-1 -0.1242 e_t-2 -0.1522 e_t-3 +  e_t

8. Checking a different model 

We would like  to check ARMA(1,1) for us to be able to compare AICs. 

```{r}
arma11<-arima(my_data,order=c(1,1,1))
arma11
```
**Model Equation:**
X_t= 0.5727 X_t-1 -0.8027 e_t-1 +  e_t

The AIC of the first model is  smaller than the AIC of the second model. Therefore, we will use the first model. 

9. Checking the model assumptions, NICE assumptions

```{r}
plot.ts(ma3$residuals)
```
The residuals has no pattern in time. 

**Test for normality**

```{r}
hist(ma3$residuals)
```

The residuals are normally distributed. 

```{r}
acf(ma3$residuals)
```

There are no spikes which means that the residuals are independent. 

**Testing for Indepenndence of Residuals**

```{r}
Box.test(ma3$residuals,lag=20,fitdf = 3)
```

Since the p value is 0.05793 which is greater than alpha (0.05), therefore we fail to reject H_0. This means residuals are independent which agrees with the ACF plot. 

10. Forecasting 

```{r}
data<-read.csv("Final Data.csv")
my_data2<-data[1:1460,2]
head(my_data2)
```

We took the first 1460 observations as our data and left 2 observations for forecasting. 

```{r}
library(forecast)
fit <- Arima(my_data2, order=c(0,1,3))
forecasts <- forecast(fit, h=2)
forecasts
```

The forecasted values for observations 1461 and 1462 are 14.61767 and 15.05462 respectively. 
The actual values for observations 1461 and 1462 are 15.05263158  and 10 respectively. 

**Calculating Forecast Error**

```{r}
forecasted_values <- c(14.61767 , 15.05462)
actual_values <-c(15.05263158, 10)
errors = abs(actual_values - forecasted_values)
errors 

```

The forecast error of the first observation is less than that of the second. This could indicate that this forecast is good for short-term forecast, because it seems that the error would increase for later observations. 


